---
title: "Power Analysis for Multivariate Analysis Using SEM"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lavaan)
```

# Power Analysis in SEM

This analysis updates the paper [A framework for power analysis using a structural equation modelling procedure](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/1471-2288-3-27)

The paper describes how to do power analysis for multivariate analysis using SEM.

The examples in the paper are written using Mx. This file updates the programs and gives some examples using R and the lavaan package. 

Many analyses are special cases of SEMs, and SEMs are pretty straightforward to calculate power for, using the [Satorra-Saris method](https://link.springer.com/article/10.1007/BF02294150).

# Examples

The basic approach is always the same. Set up the data which matches the population that we are interested in, by making a model in lavaan that is saturated. Then get the implied means and covariance matrix (or matrices). Then fit the data to the model with the parameter of interest constrained to zero.

The chi-square from this model with the parameter constrained to zero is the non-centrality parameter. 


## t test
We'll start with some simple examples, like a t-test. 

### Power from Base R

First we'll calculate power for a t-test using the power.t.test() function, for d = 0.2, n = 100 (per group).

```{r t1}

power.t.test(d = 0.2, n = 100)

```

### Power from lavaan.

Now we estimate power using lavaan.  This can be done in two ways: using a single group with a dichotous predictor, or using multiple groups. 

Single group first. We need to generate some data for lavaan with the right variables in it (what the data actually have in them doesn't really matter).

```{r t2}

d <- data.frame(x = rep(c(0, 1), 100), y = rnorm(200))

```
Now we set up our model in lavaan.

We need to set the variance of y (which we'll set to 1), the variance of x (which is determined by the values of x - the variance is calculated using $p * (1 - p)$  )

```{r t3}
t_m1 <- "
y ~ 0.2x
y ~~ 1
"

```


